---
title: "TÜİK Tourism Forecasting Project"
author: "Leyla Shikhzaman - 138723991"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Packages and Setup

```{r packages}
library(tuikr)
library(httr)
library(readxl)
library(dplyr)
library(zoo)
library(forecast)
library(ggplot2)
```

## 2. Data Access via tuikr

```{r data-access}
# Access TÜİK Tourism tables through tuikr (Theme ID 14)
# Data access date:
access_date <- Sys.Date()
cat("Data access date:", format(access_date), "\n")

# Get tourism tables
tourism_tables <- statistical_tables("14")

# Start session to access istab file
s <- handle("https://veriportali.tuik.gov.tr")
GET(handle = s, path = "/")

# Download the selected table (row 57: Number of Trips by Accommodation and Quarter)
tmp_xls <- tempfile(fileext = ".xls")
GET(tourism_tables$table_url[57],
    handle = s,
    write_disk(tmp_xls, overwrite = TRUE),
    add_headers(
      "User-Agent" = "Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36",
      "Referer" = "https://veriportali.tuik.gov.tr/"
    ))

# Read raw file
df_raw <- read_excel(tmp_xls, sheet = 1)
cat("Raw data dimensions:", nrow(df_raw), "rows,", ncol(df_raw), "columns\n")
```

## 3. Data Preparation

```{r data-prep}
# Clean and structure the data
df_clean <- df_raw %>%
  rename(year = 1, quarter = 2, total_trips = 3) %>%
  mutate(year = na.locf(year, na.rm = FALSE)) %>%
  filter(quarter %in% c("I", "II", "III", "IV")) %>%
  select(year, quarter, total_trips) %>%
  mutate(total_trips = as.numeric(gsub(",", "", total_trips))) %>%
  mutate(q_num = ifelse(quarter=="I",1,
                 ifelse(quarter=="II",2,
                 ifelse(quarter=="III",3,4)))) %>%
  mutate(period = paste0(year, " Q", q_num)) %>%
  filter(!is.na(total_trips)) %>%
  select(year, quarter, q_num, total_trips, period)

# Report key facts
cat("Latest available observation:", tail(df_clean$period, 1), "\n")
cat("Forecast target period: 2026 Q1\n")
cat("Total observations:", nrow(df_clean), "\n")
cat("Missing values:", sum(is.na(df_clean$total_trips)), "\n")

# Create time series object (quarterly, starting 2009 Q1)
ts_data <- ts(df_clean$total_trips, start = c(2009, 1), frequency = 4)
print(ts_data)
```

## 4. Exploratory Time Series Analysis

```{r eda}
# Plot the actual time series
p_actual <- ggplot(df_clean, aes(x = seq_along(total_trips), y = total_trips)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 1.5) +
  scale_x_continuous(breaks = seq(1, 68, by = 4),
                     labels = df_clean$period[seq(1, 68, by = 4)]) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Number of Domestic Trips by Quarter (2009-2025)",
       x = "Period", y = "Number of Trips (thousands)")

print(p_actual)
ggsave("outputs/figures/actual_series_plot.png", p_actual, width=10, height=5)

# Decomposition to check trend and seasonality
decomp <- decompose(ts_data)
plot(decomp)
```

## 5. Forecasting Methods

### 5.1 Naïve Forecasting

```{r naive}
naive_fit <- naive(ts_data, h = 1)
naive_fitted <- fitted(naive_fit)
naive_errors <- df_clean$total_trips[-1] - as.numeric(naive_fitted)[-1]

naive_bias  <- mean(naive_errors)
naive_mad   <- mean(abs(naive_errors))
naive_mse   <- mean(naive_errors^2)
naive_mape  <- mean(abs(naive_errors / df_clean$total_trips[-1])) * 100
naive_rsfe  <- sum(naive_errors)
naive_ts    <- naive_rsfe / naive_mad
naive_next  <- as.numeric(naive_fit$mean[1])

cat("Naive next-period forecast:", naive_next, "\n")

p_naive <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = c(NA, as.numeric(naive_fitted)[-1]), color = "Naive Forecast")) +
  scale_color_manual(values = c("Actual"="steelblue","Naive Forecast"="red")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title="Actual vs Naive Forecast", x="Period", y="Trips (thousands)", color="")

print(p_naive)
ggsave("outputs/figures/naive_forecast_plot.png", p_naive, width=10, height=5)
```

### 5.2 Moving Average

```{r moving-average}
# Window = 4 (one full year, appropriate for quarterly data)
ma_values <- ma(ts_data, order = 4)
ma_vec <- as.numeric(ma_values)
actual_vec <- as.numeric(ts_data)

ma_errors <- actual_vec[!is.na(ma_vec)] - ma_vec[!is.na(ma_vec)]
ma_bias  <- mean(ma_errors)
ma_mad   <- mean(abs(ma_errors))
ma_mse   <- mean(ma_errors^2)
ma_mape  <- mean(abs(ma_errors / actual_vec[!is.na(ma_vec)])) * 100
ma_rsfe  <- sum(ma_errors)
ma_ts    <- ma_rsfe / ma_mad
# Next period forecast = last available MA value
ma_next  <- tail(ma_vec[!is.na(ma_vec)], 1)

cat("Moving Average next-period forecast:", ma_next, "\n")

p_ma <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = ma_vec, color = "Moving Average")) +
  scale_color_manual(values = c("Actual"="steelblue","Moving Average"="orange")) +
  theme_minimal() +
  labs(title="Actual vs Moving Average (window=4)", x="Period", y="Trips (thousands)", color="")

print(p_ma)
ggsave("outputs/figures/moving_average_plot.png", p_ma, width=10, height=5)
```

### 5.3 Weighted Moving Average

```{r wma}
# Weights: more weight on recent quarters
weights <- c(0.1, 0.2, 0.3, 0.4)

wma_vec <- rep(NA, length(actual_vec))
for (i in 4:length(actual_vec)) {
  wma_vec[i] <- sum(weights * actual_vec[(i-3):i])
}

wma_errors <- actual_vec[!is.na(wma_vec)] - wma_vec[!is.na(wma_vec)]
wma_bias  <- mean(wma_errors)
wma_mad   <- mean(abs(wma_errors))
wma_mse   <- mean(wma_errors^2)
wma_mape  <- mean(abs(wma_errors / actual_vec[!is.na(wma_vec)])) * 100
wma_rsfe  <- sum(wma_errors)
wma_ts    <- wma_rsfe / wma_mad
wma_next  <- sum(weights * tail(actual_vec, 4))

cat("WMA next-period forecast:", wma_next, "\n")

p_wma <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = wma_vec, color = "WMA Forecast")) +
  scale_color_manual(values = c("Actual"="steelblue","WMA Forecast"="purple")) +
  theme_minimal() +
  labs(title="Actual vs Weighted Moving Average", x="Period", y="Trips (thousands)", color="")

print(p_wma)
ggsave("outputs/figures/weighted_moving_average_plot.png", p_wma, width=10, height=5)
```

### 5.4 Exponential Smoothing

```{r exp-smoothing}
ses_fit  <- ses(ts_data, alpha = 0.3, h = 1)
ses_fitted <- as.numeric(fitted(ses_fit))
ses_errors <- actual_vec - ses_fitted

ses_bias  <- mean(ses_errors)
ses_mad   <- mean(abs(ses_errors))
ses_mse   <- mean(ses_errors^2)
ses_mape  <- mean(abs(ses_errors / actual_vec)) * 100
ses_rsfe  <- sum(ses_errors)
ses_ts    <- ses_rsfe / ses_mad
ses_next  <- as.numeric(ses_fit$mean[1])

cat("Exponential Smoothing next-period forecast:", ses_next, "\n")

p_ses <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = ses_fitted, color = "Exp Smoothing")) +
  scale_color_manual(values = c("Actual"="steelblue","Exp Smoothing"="darkgreen")) +
  theme_minimal() +
  labs(title="Actual vs Exponential Smoothing (alpha=0.3)", x="Period", y="Trips (thousands)", color="")

print(p_ses)
ggsave("outputs/figures/exponential_smoothing_plot.png", p_ses, width=10, height=5)
```

### 5.5 Trend-Adjusted Exponential Smoothing

```{r holt}
holt_fit <- holt(ts_data, alpha = 0.3, beta = 0.1, h = 1)
holt_fitted <- as.numeric(fitted(holt_fit))
holt_errors <- actual_vec - holt_fitted

holt_bias  <- mean(holt_errors)
holt_mad   <- mean(abs(holt_errors))
holt_mse   <- mean(holt_errors^2)
holt_mape  <- mean(abs(holt_errors / actual_vec)) * 100
holt_rsfe  <- sum(holt_errors)
holt_ts    <- holt_rsfe / holt_mad
holt_next  <- as.numeric(holt_fit$mean[1])

cat("Holt next-period forecast:", holt_next, "\n")

p_holt <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = holt_fitted, color = "Holt Forecast")) +
  scale_color_manual(values = c("Actual"="steelblue","Holt Forecast"="brown")) +
  theme_minimal() +
  labs(title="Actual vs Trend-Adjusted Exponential Smoothing", x="Period", y="Trips (thousands)", color="")

print(p_holt)
ggsave("outputs/figures/trend_adjusted_smoothing_plot.png", p_holt, width=10, height=5)
```

### 5.6 Linear Trend Projection

```{r linear-trend}
t_index <- 1:nrow(df_clean)
lm_fit <- lm(total_trips ~ t_index, data = df_clean)
summary(lm_fit)

lm_fitted <- as.numeric(fitted(lm_fit))
lm_errors <- df_clean$total_trips - lm_fitted

lm_bias  <- mean(lm_errors)
lm_mad   <- mean(abs(lm_errors))
lm_mse   <- mean(lm_errors^2)
lm_mape  <- mean(abs(lm_errors / df_clean$total_trips)) * 100
lm_rsfe  <- sum(lm_errors)
lm_ts    <- lm_rsfe / lm_mad
lm_next  <- predict(lm_fit, newdata = data.frame(t_index = nrow(df_clean) + 1))

cat("Linear Trend next-period forecast:", lm_next, "\n")

p_lm <- ggplot(df_clean, aes(x = t_index)) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = lm_fitted, color = "Linear Trend")) +
  scale_color_manual(values = c("Actual"="steelblue","Linear Trend"="red")) +
  theme_minimal() +
  labs(title="Actual vs Linear Trend Projection", x="Period", y="Trips (thousands)", color="")

print(p_lm)
ggsave("outputs/figures/trend_projection_plot.png", p_lm, width=10, height=5)
```

### 5.7 Seasonal Indices

```{r seasonal-indices}
# Calculate average for each quarter
seasonal_avg <- df_clean %>%
  group_by(q_num) %>%
  summarise(avg = mean(total_trips))

overall_avg <- mean(df_clean$total_trips)
seasonal_avg$index <- seasonal_avg$avg / overall_avg

cat("Seasonal Indices:\n")
print(seasonal_avg)
cat("Q1 (lowest):", round(seasonal_avg$index[1],3), "\n")
cat("Q3 (highest):", round(seasonal_avg$index[3],3), "\n")

# Deseasonalized values
df_clean$seas_index <- seasonal_avg$index[df_clean$q_num]
df_clean$deseason   <- df_clean$total_trips / df_clean$seas_index

# Trend on deseasonalized
t_index2 <- 1:nrow(df_clean)
lm_deseas <- lm(deseason ~ t_index2, data = df_clean)
deseas_trend <- predict(lm_deseas, newdata = data.frame(t_index2 = nrow(df_clean) + 1))
si_next <- deseas_trend * seasonal_avg$index[1]  # Q1 for 2026

si_fitted <- fitted(lm_deseas) * df_clean$seas_index
si_errors <- df_clean$total_trips - si_fitted
si_bias  <- mean(si_errors)
si_mad   <- mean(abs(si_errors))
si_mse   <- mean(si_errors^2)
si_mape  <- mean(abs(si_errors / df_clean$total_trips)) * 100
si_rsfe  <- sum(si_errors)
si_ts    <- si_rsfe / si_mad

cat("Seasonal Indices next-period forecast:", si_next, "\n")

p_si <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = si_fitted, color = "Seasonal Indices")) +
  scale_color_manual(values = c("Actual"="steelblue","Seasonal Indices"="darkgreen")) +
  theme_minimal() +
  labs(title="Actual vs Seasonal Indices Forecast", x="Period", y="Trips (thousands)", color="")

print(p_si)
ggsave("outputs/figures/seasonal_indices_plot.png", p_si, width=10, height=5)
```

### 5.8 Additive Decomposition

```{r additive-decomp}
decomp_add <- decompose(ts_data, type = "additive")
plot(decomp_add)

add_fitted <- ts_data - decomp_add$random
add_errors <- as.numeric(ts_data) - as.numeric(add_fitted)
add_errors <- add_errors[!is.na(add_errors)]
actual_noNA <- as.numeric(ts_data)[!is.na(as.numeric(add_fitted))]

add_bias  <- mean(add_errors)
add_mad   <- mean(abs(add_errors))
add_mse   <- mean(add_errors^2)
add_mape  <- mean(abs(add_errors / actual_noNA)) * 100
add_rsfe  <- sum(add_errors)
add_ts    <- add_rsfe / add_mad

# Forecast next period using trend + seasonal
trend_val <- tail(as.numeric(decomp_add$trend)[!is.na(decomp_add$trend)], 1)
seas_val  <- as.numeric(decomp_add$seasonal)[1]  # Q1 seasonal
add_next  <- trend_val + seas_val

cat("Additive Decomposition next-period forecast:", add_next, "\n")

add_fitted_vec <- as.numeric(add_fitted)
p_add <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = add_fitted_vec, color = "Additive Decomp")) +
  scale_color_manual(values = c("Actual"="steelblue","Additive Decomp"="orange")) +
  theme_minimal() +
  labs(title="Actual vs Additive Decomposition", x="Period", y="Trips (thousands)", color="")

print(p_add)
ggsave("outputs/figures/additive_decomposition_plot.png", p_add, width=10, height=5)
```

### 5.9 Multiplicative Decomposition

```{r mult-decomp}
decomp_mult <- decompose(ts_data, type = "multiplicative")

mult_fitted <- ts_data / decomp_mult$random
mult_errors <- as.numeric(ts_data) - as.numeric(mult_fitted)
mult_errors <- mult_errors[!is.na(mult_errors)]
actual_noNA2 <- as.numeric(ts_data)[!is.na(as.numeric(mult_fitted))]

mult_bias  <- mean(mult_errors)
mult_mad   <- mean(abs(mult_errors))
mult_mse   <- mean(mult_errors^2)
mult_mape  <- mean(abs(mult_errors / actual_noNA2)) * 100
mult_rsfe  <- sum(mult_errors)
mult_ts    <- mult_rsfe / mult_mad

trend_val2 <- tail(as.numeric(decomp_mult$trend)[!is.na(decomp_mult$trend)], 1)
seas_val2  <- as.numeric(decomp_mult$seasonal)[1]
mult_next  <- trend_val2 * seas_val2

cat("Multiplicative Decomposition next-period forecast:", mult_next, "\n")

mult_fitted_vec <- as.numeric(mult_fitted)
p_mult <- ggplot(df_clean, aes(x = seq_along(total_trips))) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = mult_fitted_vec, color = "Mult Decomp")) +
  scale_color_manual(values = c("Actual"="steelblue","Mult Decomp"="purple")) +
  theme_minimal() +
  labs(title="Actual vs Multiplicative Decomposition", x="Period", y="Trips (thousands)", color="")

print(p_mult)
ggsave("outputs/figures/multiplicative_decomposition_plot.png", p_mult, width=10, height=5)
```

### 5.10 Regression with Trend and Seasonal Dummies

```{r regression-dummies}
# Create seasonal dummy variables (Q1 is reference)
df_clean$Q2 <- ifelse(df_clean$q_num == 2, 1, 0)
df_clean$Q3 <- ifelse(df_clean$q_num == 3, 1, 0)
df_clean$Q4 <- ifelse(df_clean$q_num == 4, 1, 0)
df_clean$t  <- 1:nrow(df_clean)

reg_fit <- lm(total_trips ~ t + Q2 + Q3 + Q4, data = df_clean)
summary(reg_fit)

reg_fitted <- as.numeric(fitted(reg_fit))
reg_errors <- df_clean$total_trips - reg_fitted

reg_bias  <- mean(reg_errors)
reg_mad   <- mean(abs(reg_errors))
reg_mse   <- mean(reg_errors^2)
reg_mape  <- mean(abs(reg_errors / df_clean$total_trips)) * 100
reg_rsfe  <- sum(reg_errors)
reg_ts    <- reg_rsfe / reg_mad
reg_next  <- predict(reg_fit, newdata = data.frame(t = nrow(df_clean)+1, Q2=0, Q3=0, Q4=0))

cat("Regression next-period forecast:", reg_next, "\n")

p_reg <- ggplot(df_clean, aes(x = t)) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = reg_fitted, color = "Regression")) +
  scale_color_manual(values = c("Actual"="steelblue","Regression"="darkred")) +
  theme_minimal() +
  labs(title="Actual vs Regression with Trend & Seasonal Dummies", x="Period", y="Trips (thousands)", color="")

print(p_reg)
ggsave("outputs/figures/regression_seasonal_dummy_plot.png", p_reg, width=10, height=5)
```

## 6. Accuracy Comparison Table

```{r accuracy-table}
accuracy_table <- data.frame(
  Method = c("Naive","Moving Average","Weighted MA","Exp Smoothing",
             "Trend-Adj ES","Linear Trend","Seasonal Indices",
             "Additive Decomp","Mult Decomp","Regression"),
  Bias  = round(c(naive_bias, ma_bias, wma_bias, ses_bias,
                  holt_bias, lm_bias, si_bias, add_bias, mult_bias, reg_bias), 2),
  MAD   = round(c(naive_mad, ma_mad, wma_mad, ses_mad,
                  holt_mad, lm_mad, si_mad, add_mad, mult_mad, reg_mad), 2),
  MSE   = round(c(naive_mse, ma_mse, wma_mse, ses_mse,
                  holt_mse, lm_mse, si_mse, add_mse, mult_mse, reg_mse), 2),
  MAPE  = round(c(naive_mape, ma_mape, wma_mape, ses_mape,
                  holt_mape, lm_mape, si_mape, add_mape, mult_mape, reg_mape), 2),
  RSFE  = round(c(naive_rsfe, ma_rsfe, wma_rsfe, ses_rsfe,
                  holt_rsfe, lm_rsfe, si_rsfe, add_rsfe, mult_rsfe, reg_rsfe), 2),
  Tracking_Signal = round(c(naive_ts, ma_ts, wma_ts, ses_ts,
                             holt_ts, lm_ts, si_ts, add_ts, mult_ts, reg_ts), 2),
  Next_Forecast = round(c(naive_next, ma_next, wma_next, ses_next,
                           holt_next, lm_next, si_next, add_next, mult_next, reg_next), 2)
)

print(accuracy_table)
write.csv(accuracy_table, "outputs/tables/accuracy_comparison.csv", row.names = FALSE)
```

## 7. Superior Method Selection

```{r superior-method}
# Based on lowest MAPE and suitability to quarterly seasonal data
# Regression with trend and seasonal dummies captures both trend and seasonality

best_method <- "Regression with Trend and Seasonal Dummy Variables"
best_forecast <- reg_next

cat("Superior Method:", best_method, "\n")
cat("Forecast for 2026 Q1:", round(best_forecast, 2), "thousand trips\n")

# Final plot
p_final <- ggplot(df_clean, aes(x = t)) +
  geom_line(aes(y = total_trips, color = "Actual")) +
  geom_line(aes(y = reg_fitted, color = "Fitted")) +
  geom_point(aes(x = nrow(df_clean)+1, y = best_forecast, color = "2026 Q1 Forecast"),
             size = 4, shape = 17) +
  scale_color_manual(values = c("Actual"="steelblue","Fitted"="darkred","2026 Q1 Forecast"="green3")) +
  theme_minimal() +
  labs(title="Superior Method: Regression with Trend & Seasonal Dummies — Forecast for 2026 Q1",
       x="Period", y="Trips (thousands)", color="")

print(p_final)
ggsave("outputs/figures/superior_method_plot.png", p_final, width=10, height=5)
```

## 8. Final Forecast

```{r final-forecast}
final_forecast <- data.frame(
  Superior_Method       = best_method,
  Data_Access_Date      = format(access_date),
  Latest_Observation    = "2025 Q4",
  Forecast_Target       = "2026 Q1",
  Forecasted_Value      = round(best_forecast, 2),
  MAPE                  = round(reg_mape, 2)
)

print(final_forecast)
write.csv(final_forecast, "outputs/tables/final_forecast.csv", row.names = FALSE)
```

## 9. Limitations

The analysis has several limitations. First, the 2020 COVID-19 pandemic caused a structural break in domestic tourism, with Q2 2020 recording the lowest value in the entire series. This outlier may distort trend and seasonal estimates. Second, the regression model assumes a linear trend, which may not hold in future periods. Third, the model does not account for external factors such as economic conditions, fuel prices, or policy changes that could affect domestic travel behavior.

## 10. Reproducibility Notes

Data was accessed directly from the TÜİK Data Portal using the tuikr R package (statistical_tables function, Theme ID 14) combined with httr::GET() to download the istab file. No data was manually downloaded, edited, or saved outside of R. To reproduce, clone the repository, open forecasting_project.Rmd in RStudio, and click Knit.